Bricking up backdoors in Intrusion Detection Systems

### 2019-09-12 paper/paper.tex
Modified paper

\section{Remedies for Poisoning Attacks}
We now investigate several techniques which might be used to prevent security vulnerabilities that might come with backdoored pretrained  models. In this respect the ability to detect a backdoor and the ability to remove a backdoor from the trained model can be considered as equally effective since one often has the option to fall back to a model obtained from a different source in case a model looks suspicious.

@@

%Several graphs have been proposed for visualizing feature dependence of non-interpretable machine learning models~\cite{goldstein2015peeking, friedman_greedy_2001, apley2016visualizing}.
%In this section, we introduce the graphs which we will use for interpreting our models' outcomes which are not available in real-world situations.
\subsubsection{Partial Dependence Plots}
\glspl{pdp} were proposed in \cite{friedman_greedy_2001} and visualize dependence of a model's predictions by plotting the \gls{mui}'s prediction for a modified dataset for which the feature's value has been fixed to a certain value and computing the an average over the dataset.

If we denote bw $\boldsymbol X \in \mathbb R ^n$ a random vector drawn from the feature space and bw $f(\boldsymbol X) \in [0,1]$ the  prediction function, the \gls{pdp} for the $i$th feature $X_i$ can be expressed as
\begin{equation}
\text{PDP}_i(w) = \mathbb E_{\boldsymbol X}\Big(f(X_1,\ldots,X_{i-1},w,X_{i+1},\ldots X_n)\Big) . % \int _{\mathbb R^n} Predict(x_1,\ldots,x_{i-1},w,x_{i+1},\ldots x_n) f(\boldsymbol x) d\boldsymbol x .
\end{equation}
Empirically, we can approximate the distribution of the feature space using the distribution of observed samples. Hence, at a given point $w,$ the \gls{pdp} can be found by setting the corresponding value of all samples in the dataset to $w$ and averaging over the predictions of the resulting modified dataset.

\subsubsection{Accumulated Local Effects}
Due to feature dependence it is very likely that in the feature space areas exist which have a very low probability to occur. Since a model is trained with real, observed data, the training set therefore does not include samples for these areas, which causes the model's predictions to become indeterminate for these areas. This poses a problem when considering these predictions for computing \glspl{pdp}.

In an attempt to overcome this problem, it is possible to only consider samples which are likely to occur for certain feature values, i.e. to consider the conditional distribution of remaining features, for computing explainability graphs. This results the concept for \gls{ale} plots.

@@

To combat ambiguity of this definition, we force ALE$_i(x)$ to have zero mean on the domain of $X_i$.

For empirical evaluation, we approximate the conditional distributions of $\boldsymbol X$ by averaging over samples for which $X_i \approx w$. We used the 10  closest samples
%to the interval's center
for estimating the distribution.

@@

\subsection{\gls{dl} defenses}

@@

To perform pruning, a validation dataset is needed. We take a validation dataset that is $\frac{1}{4}$ of the training dataset. We use the validation dataset for pruning as described in the next sections and use a test dataset that is also $\frac{1}{4}$ to verify whether the backdoor can successfully be removed and how much the accuracy on the original data suffers. Training, validation and test dataset are pairwise disjoint.

@@

% AH. moved 'by their average activation' to beginning of sentence to make clear it applies to both variants
We implement two variants of the pruning defense \cite{gu_badnets:_2017}: Pruning neurons by their average activation in \begin{enumerate*}\item all layers \item only in the last layer \end{enumerate*}. Neurons with the smallest average activation (the least used ones) are pruned first. For this purpose, we look at the value of the activation function (\gls{relu} in our case) in the corresponding layer for all samples in the validation dataset and prune neurons by setting the weight and bias to 0 in the layer preceding the activation function.

@@

In addition to fine-tuning we also tried fine-pruning~\cite{liu_fine-pruning:_2018}, i.e. we pruned 50\% of neural network neurons and applied fine-tuning to the resulting network. However, also fine-pruning did not result in convenient backdoor removal performance.

The use of fine-tuning exclusively makes  sense if both the computational effort and the required training set size for fine-tuning are substantially lower than for the original training procedure. Hence, a decay of the backdoor efficacy as depicted in \autoref{fig:finetuning} cannot be considered sufficient.

@@

%We propose a defense approach that can be used for similar use cases like the mechanisms recently proposed to remove backdoors from \glspl{cnn}, discussed above. 
The concept behind our pruning defense is that leaves that are never used by samples in the validation dataset might be used by backdoored inputs. If these ``useless'' leaves are removed, performance of the classifier on the validation dataset should not decrease
%by a lot
significantly,
while decisions that are used for the backdoor are likely to be removed.
We developed several variants of our pruning defense with increasing levels of sophistication:
\begin{enumerate}[wide, labelwidth=!, labelindent=0pt]
\item Pruning leaves based on their usage: This means that the least used leaf is pruned first and the most used one last.
\item Like (1) but consider only ``harmless'' leaves: We assume that attackers want malicious samples appear harmless.
\item Like (2) but use depth to decide when two leaves used equally often: The rationale is that we assume that hand-crafted backdoors require fewer rules to classify than regular samples and are thus have lower depth.
\end{enumerate}

\autoref{fig:pruningExample} shows an example of variant (3) of pruning applied to a tree
\begin{figure}[h]
\includegraphics[width=\columnwidth]{../prune_CAIA_backdoor_15/prune_oh_d.pdf}
\caption{Pruning an \gls{rf} for \unsw{}.}
\label{fig:performancePruningRF}
\end{figure}
%\todo{Max: Which pruning algorithm actually?}
and \autoref{fig:performancePruningRF} shows the resulting accuracy for pruning according to variant (3) for the \unsw{} dataset. With sufficient leaves being pruned, the accuracy of the backdoor approaches zero. The accuracy for the regular data does not decrease, it even might even increase by
%thanks to the the pruning also
reducing overfitting. %With a significantly smaller validation set, pruning takes longer to remove the backdoor
%until it works
%but it still works. This is an important finding as a large validation dataset makes it hard for users without access to expensive hardware to perform the pruning.
Even with a very small validation dataset of just 1\% of its original size, the pruning still works as expected.

For \cic{} we get similar results, but the pruning does not remove the backdoor completely ($\sim$30\% backdoor accuracy remains). We attribute this to the fact, that in this dataset there are also regular flows which have $\text{stddev}(\text{TTL}) > 0$. Thus, backdoored flows cannot always be sharply distinguished from regular flows. We find that pruning only harmless leaves is very beneficial for \unsw{} but not for \cic{}. However, considering depth when making the decision which leaf to prune first, leads to the backdoor being removed considerably earlier in the pruning process.

@@

To ensure that no backdoor is contained in the model in question, it has to be carefully analyzed for questionable decisions and potentially unnecessary features. For this purpose, \gls{pdp} and \gls{ale} plots are an effective tool. In fact, already throughout training process explainability plots constitute a useful tool to ensure that the model is not unintentionally trained  to artifacts the dataset yields. Indeed, the implementation of a backdoor as conducted in this research is only possible when using several features involving the \gls{ttl} value. Even though it might seem tempting to provide all possible features to a \gls{dl} or \gls{rf} classifier and let it learn the most important ones, this strategy should be avoided.

>>>>

\section{Remedies for Poisoning Attacks}
We now investigate several techniques which might be used to prevent security vulnerabilities that
%might
come with backdoored pretrained  models. In this respect, the ability to detect a backdoor and the ability to remove a backdoor from the trained model can be considered as equally effective since one often has the option to fall back to a model obtained from a different source in case a model looks suspicious.

@@

%Several graphs have been proposed for visualizing feature dependence of non-interpretable machine learning models~\cite{goldstein2015peeking, friedman_greedy_2001, apley2016visualizing}.
%In this section, we introduce the graphs which we will use for interpreting our models' outcomes which are not available in real-world situations.
\subsubsection{Partial Dependence Plots}
\glspl{pdp} were proposed in \cite{friedman_greedy_2001} and visualize dependence of a model's predictions by plotting the \gls{mui}'s prediction for a modified dataset for which the feature's value has been fixed to a certain value, averaging over the modified dataset.

If we denote by $\boldsymbol X \in \mathbb R ^n$ a random vector drawn from the feature space and by $f(\boldsymbol X) \in [0,1]$ the  prediction function, the \gls{pdp} for the $i$th feature $X_i$ can be expressed as
\begin{equation}
\text{PDP}_i(w) = \mathbb E_{\boldsymbol X}\Big(f(X_1,\ldots,X_{i-1},w,X_{i+1},\ldots X_n)\Big) . % \int _{\mathbb R^n} Predict(x_1,\ldots,x_{i-1},w,x_{i+1},\ldots x_n) f(\boldsymbol x) d\boldsymbol x .
\end{equation}
Empirically, we can approximate the distribution of the feature space using the distribution of observed samples. Hence, at a given point $w,$ the \gls{pdp} can be found by setting the corresponding value of all samples in the dataset to $w$ and averaging over the predictions of the resulting modified dataset.

\subsubsection{Accumulated Local Effects}
Due to feature dependence it is very likely that areas exist in the feature space which are very unlikely
% have a very low probability
to occur. Since a model is trained with real, observed data, the training set therefore does not include samples for these areas, which causes the model's predictions to become indeterminate for these areas. This poses a problem when considering these predictions for computing \glspl{pdp}.

In an attempt to overcome this problem, it is possible to only consider samples which are likely to occur for certain feature values, i.e. to consider the conditional distribution of remaining features, for computing explainability graphs. This results in the concept of \gls{ale} plots.

@@

To combat ambiguity of this definition, we force ALE$_i(x)$ to have zero mean on the domain of $X_i$.
For empirical evaluation, we approximate the conditional distributions of $\boldsymbol X$ by averaging over samples for which $X_i \approx w$. In this paper, we used the 10  closest samples
%to the interval's center
for estimating the distribution.

@@

\subsection{\gls{dl} Poisoning Defenses}

@@

To perform pruning, a validation dataset is needed. We take a validation dataset that is $\frac{1}{4}$ of the training dataset. We use the validation dataset for pruning as described in the next sections and a test dataset that is also $\frac{1}{4}$ to verify whether the backdoor can successfully be removed and how much the accuracy on the original data suffers. Training, validation and test dataset are pairwise disjoint.

@@

% AH. moved 'by their average activation' to beginning of sentence to make clear it applies to both variants
We implement two variants of the pruning defense \cite{gu_badnets:_2017}: Pruning neurons by their average activation in \begin{enumerate*}\item all layers and \item only in the last layer \end{enumerate*}. Neurons with the smallest average activation (the least used ones) are pruned first. For this purpose, we look at the value of the activation function (\gls{relu} in our case) in the corresponding layer for all samples in the validation dataset and prune neurons by setting the weight and bias to 0 in the layer preceding the activation function.

@@

In addition to fine-tuning we also tried fine-pruning~\cite{liu_fine-pruning:_2018}, preceding fine-tuning with a large number of different pruning strategies. From these experiments, only pruning a certain fraction of the \gls{mlp}'s first layer resulted in a discernible drop of backdoor efficacy after one epoch of fine-tuning, primarily for \cic{}.

The significance of the \gls{mlp}'s first layer for the backdoor presumably results from the simplicity of our backdoor pattern. Hence, applicability of this pruning technique for general situations is questionable. We conclude that also fine-pruning cannot be considered a reliable method for backdoor removal in the context of \glspl{ids}.

%The use of fine-tuning exclusively makes  sense if both the computational effort and the required training set size for fine-tuning are substantially lower than for the original training procedure. Hence, a decay of the backdoor efficacy as depicted in \autoref{fig:finetuning} cannot be considered sufficient.

@@

%We propose a defense approach that can be used for similar use cases like the mechanisms recently proposed to remove backdoors from \glspl{cnn}, discussed above. 
The concept behind our pruning defense is that leaves that are never used by samples in the validation dataset might be used by backdoored inputs. If these ``useless'' leaves are removed, performance of the classifier on the validation dataset should not decrease
%by a lot
significantly,
while decisions that are used for the backdoor are likely to be removed.
We developed several variants of our pruning defense with increasing levels of sophistication:
\begin{enumerate}[wide, labelwidth=!, labelindent=0pt]
\item Pruning leaves based on their usage: This means that the least used leaf is pruned first and the most used one last.
\item Like (1) but consider only ``harmless'' leaves: We assume that attackers want malicious samples appear harmless.
\item Like (2) but additionally use depth to decide when
%two % AH could be more than two
leaves are used equally often: The rationale is that we assume that hand-crafted backdoors require fewer rules to classify than regular samples and thus have lower depth.
\end{enumerate}

\autoref{fig:pruningExample} shows an example of variant (3) of pruning applied to a tree
\begin{figure}[h]
\includegraphics[width=\columnwidth]{../prune_CAIA_backdoor_15/prune_oh_d.pdf}
\caption{Pruning an \gls{rf} for \unsw{}.}
\label{fig:performancePruningRF}
\end{figure}
%\todo{Max: Which pruning algorithm actually?}
and \autoref{fig:performancePruningRF} shows the resulting accuracy for pruning according to variant (3) for the \unsw{} dataset. With sufficient leaves being pruned, the accuracy of the backdoor approaches zero. The accuracy for the regular data does not decrease, it might even increase by
%thanks to the the pruning also
reducing overfitting. %With a significantly smaller validation set, pruning takes longer to remove the backdoor
%until it works
%but it still works. This is an important finding as a large validation dataset makes it hard for users without access to expensive hardware to perform the pruning.
Even with a very small validation dataset of just 1\% of its original size, the pruning still works as expected.

For \cic{} we get similar results, but the pruning does not remove the backdoor completely ($\sim$30\% backdoor accuracy remains). We attribute this to the fact, that in this dataset there are also regular flows which have $\text{stdev}(\text{TTL}) > 0$. Thus, backdoored flows cannot always be sharply distinguished from regular flows. We find that pruning only harmless leaves is very beneficial for \unsw{} but not for \cic{}. However, considering depth when making the decision which leaf to prune first, leads to the backdoor being removed considerably earlier in the pruning process.

@@

To ensure that no backdoor is contained in the model,
%in question
it has to be  analyzed carefully for questionable decisions and potentially unnecessary features. For this purpose, \gls{pdp} and \gls{ale} plots are an effective tool. In fact, already throughout the training process explainability plots constitute a useful tool to ensure that the model is not unintentionally trained  to artifacts the dataset yields.
Indeed, the implementation of a backdoor as conducted in this research is only possible when using several features involving the \gls{ttl} value. Even though it might seem tempting to provide all possible features to a \gls{dl} or \gls{rf} classifier and let it learn the most important ones, this strategy should be avoided.
